---
title: "q3"
author: "ML"
date: "11/2/2021"
output: html_document
---

# Housing prices and log transformations

```{r libraries, message=FALSE}

library(tidyverse)
library(broom)
library(patchwork)
library(performance)

library(sjPlot)
library(sjmisc)
library(sjlabelled)
```


## Dataset

```{r df, message=FALSE}

housing <- read_csv("data/kingCountyHouses.csv")
head(housing, 10)
```

## GLM: A linear model to predict a house’s selling price

```{r explorePlot}

df1 <- housing %>% 
  select("price", "sqft") 
summary(df1)

## EDA graph

price_hist <- ggplot(data = housing, aes(x = price)) + 
  geom_histogram(binwidth = 200000L, fill = "white",
                 color = "black") + 
  xlab("Selling price of the house") + 
  ylab("Frequency") + 
  labs(title="(a)")

sqft_hist <- ggplot(data = housing, aes(x = sqft)) + 
  geom_histogram(binwidth = 500, fill = "white",
                 color = "black") + 
  xlab("Interior square footage") + 
  ylab("Frequency") + 
  labs(title="(b)")

price_hist|sqft_hist

plot(housing$price, housing$sqft)

```
Note: The selling price and square footage are both positively-skewed. 

## Model 1: Simple linear regression model

```{r m1}

## Price as the function of sqft

m1 <- lm(price ~ sqft, data = housing)
tidy(summary(m1))

check_model(m1)

```

There are violation in the assumptions. 
1. I: There is no independence in the price. Houses with larger square feet are always sold at a higher price. 
2. N: Data is not normal. It is positively-skewed. 
3. E: Price are not distributed at each level of house square feet. 


The slope coefficient \beta_{1} is 280.624. The price increases by 280.624 dollars for each square feet increase. 


## Model 2: Log transformation of Price

```{r m2}

housing <- housing %>% 
  mutate(logprice = log(price))

m2 <- lm(logprice ~ sqft, data = housing)
tidy(summary(m2))

check_model(m2)
plot(housing$sqft, housing$logprice)

compare_performance(m1, m2)

```

The price increases by 3.987e-04 with each 1 square feet increase. 

## Model 3: Log transformation of sqft

```{r m3}

housing <- housing %>% 
  mutate(logsqft = log(sqft))

m3 <- lm(price ~ logsqft, data = housing)
tidy(summary(m3))
check_model(m3)

plot(housing$logsqft, housing$price)

```



## Model 4: Log transformation of both variables

```{r m4}

m4 <- lm(logprice ~ logsqft, data = housing)
tidy(summary(m4))
check_model(m1)

plot(housing$logsqft, housing$logprice)

compare_performance(m1, m2, m3, m4, rank = TRUE)

```

# Summary

```{r finalnote}

tab_model(m1, m2)

tab_model(m3, m4)
```

