---
title: "Bank salary"
author: "ST"
date: "11/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. **Gender discrimination in bank salaries**.  In the 1970’s, Harris Trust was sued for gender discrimination in the salaries it paid its employees.  One approach to addressing this issue was to examine the starting salaries of all skilled, entry-level clerical workers between 1965 and 1975.  The following variables, which can be found in `banksalary.csv`, were collected for each worker [@Ramsey2002]:  
    - `bsal` = beginning salary (annual salary at time of hire)
    - `sal77` = annual salary in 1977
    - `sex` = MALE or FEMALE
    - `senior` = months since hired
    - `age` = age in months
    - `educ` = years of education
    - `exper` = months of prior work experience </ul>
    Creating an indicator variable based on `sex` could be helpful.  
    ```{r}
library(knitr) 
library(gridExtra)
library(kableExtra)
library(jtools)
library(GGally)
library(rsample)
library(broom)
library(tidyverse)  
banksalary <- read_csv("data\banksalary.csv")
```
    a. Identify observational units, the response variable, and explanatory variables.
- Observational units: 93 observation units.
- Response variable: beginning salary (annual salary at time of hire)
- Explanatory variables: sex, sal77, senior, age, educ, exper
    
    b. The mean starting salary of male workers (\$5957) was 16\% higher than the mean starting salary of female workers (\$5139).  Confirm these mean salaries.  Is this enough evidence to conclude gender discrimination exists?  If not, what further evidence would you need?
```{r}
meanbsal <- banksalary %>% group_by(sex) %>% summarise(mean=mean(bsal))
```
- Not enough evidence. Evidence from statistical analysis needed.

    c. How would you expect age, experience, and education to be related to starting salary?  Generate appropriate exploratory plots; are the relationships as you expected?  What implications does this have for modeling?
```{r}
ggpairs(data = banksalary,
columns = c("age", "educ", "exper", "bsal"))
```
- Age, experience and education are expected to correlate with starting salary. However, only education is correlated with starting salary, according to the plot. 
- Implication: Education should be taken into account when examining gender effect on starting salary. 
    
    
    d. Why might it be important to control for seniority (number of years with the bank) if we are only concerned with the salary when the worker started?
    - The year of employment might affect the starting salary, and everyone has different starting point. 
    
    e. By referring to exploratory plots and summary statistics, are any explanatory variables (including sex) closely related to each other?  What implications does this have for modeling?
```{r}
ggpairs(data = banksalary,
columns = c("sex","age", "educ", "exper", "bsal", "senior"))
```
    - Age and experience are closely related. We might consider including only one of the variables in the model.
    
    f. Fit a simple linear regression model with starting salary as the response and experience as the sole explanatory variable (Model 1).  Interpret the intercept and slope of this model; also interpret the R-squared value.  Is there a significant relationship between experience and starting salary?
```{r}
model_bank <- lm(bsal ~ exper, data = banksalary)
summary(model_bank)
```
    -The model estimates that when experience is 0, the starting salary would be 5289.
    -The model estimates that when starting salary increases by 1.30 with every unit increase in experience.
    -Experience explained 1.71% of the variance in starting salary.
    -No significant relationship between experience and starting salary.
    
    g. Does Model 1 meet all linear least squares regression assumptions?  List each assumption and how you decided if it was met or not.
```{r}
library(performance)
check_model(model_bank)
```
- L: Linearity violated.
- I: Not violated. 
- N: Not violated.
- E: Violated.

    h. Is a model with all 4 confounding variables (Model 2, with `senior`, `educ`, `exper`, and `age`) better than a model with just experience (Model 1)?  Justify with an appropriate significance test in addition to summary statistics of model performance.
```{r}
model_bank2 <- lm(bsal ~ senior + educ + exper + age, data = banksalary)
summary(model_bank2)

anova(model_bank, model_bank2)
```
    - Yes, model 2 is better than model 1.
    
    i. You should have noticed that the term for age was not significant in Model 2.  What does this imply about age and about future modeling steps?
    - We can exclude age from future models.
    
    j. Generate an appropriate coded scatterplot to examine a potential age-by-experience interaction.  How would you describe the nature of this interaction?
```{r}
ggplot(banksalary, aes(x = exper, y = bsal)) +
geom_point(aes(color = age)) +
scale_color_gradientn(colors = c("#00AFBB", "#E7B800", "#FC4E07"))

ggplot(banksalary, aes(x = exper, y = age)) + geom_point() 
```
- When age increases, experience increases.
    
    k. A potential final model (Model 3) would contain terms for seniority, education, and experience in addition to sex.  Does this model meet all regression assumptions?  State a 95\% confidence interval for sex and interpret this interval carefully in the context of the problem.
```{r}
model_bank3 <- lm(bsal ~ senior + educ + exper + sex, data = banksalary)
summary(model_bank3)

check_model(model_bank3)
```
    - Yes, this model appears to fulfill all the assumptions.
```{r}
confint(model_bank3)
```
    - We can be 95% confident that the starting salary of male is between 488.15 and 956.45 higher than female, after accounting for seniority, experience and education.
    
    l. Based on Model 3, what conclusions can be drawn about gender discrimination at Harris Trust?  Do these conclusions have to be qualified at all, or are they pretty clear cut?
    - There is a gender discrimination in starting salary at Harris Trust, after accounting for seniority, experience and education.
    
    m. Often salary data is logged before analysis.  Would you recommend logging starting salary in this study?  Support your decision analytically.
```{r}
ggplot(banksalary,aes(x=bsal))+
  geom_histogram()

ggplot(banksalary,aes(x=log(bsal)))+
  geom_histogram()
```
    - The distribution of raw starting salary is positively skewed. We recommend logging the data in order to achieve normal distribution.
    
    n. Regardless of your answer to the previous question, provide an interpretation for the coefficient for the male coefficient in a modified Model 3 after logging starting salary.
```{r}
model_bank4 <- lm(log(bsal) ~ senior + educ + exper + sex, data = banksalary)
summary(model_bank4)

check_model(model_bank4)
```
    - This model estimates that the starting salary of male are log(0.1295564) higher than female after accounting for seniority, experience and education.
    
    o. Build your own final model for this study and justify the selection of your final model.  You might consider interactions with gender, since those terms could show that discrimination is stronger among certain workers.  Based on your final model, do you find evidence of gender discrimination at Harris Trust?  
```{r}
final_model1 <- lm(log(bsal) ~ senior + educ*sex + exper*sex, data = banksalary)
summary(final_model1)

final_model2 <- lm(log(bsal) ~ senior + educ + exper*sex, data = banksalary)
summary(final_model2)

final_model3 <- lm(log(bsal) ~ senior + educ*sex + exper, data = banksalary)
summary(final_model3)

final_model4 <- lm(log(bsal) ~ senior*sex + educ*sex + exper*sex, data = banksalary)
summary(final_model4)

compare_performance(final_model1,final_model2,final_model3,final_model4)
anova(final_model1,final_model2)
anova(final_model1,final_model3)
anova(final_model1,final_model4)
```
    -According to AIC and BIC, final_model1 is the best model. Based on this model (final_model1), we don't find evidence of gender discrimination after accounting for seniority, interaction between experience and sex, as well as interaction between education and sex.